A thermodynamic view of dusty protoplanetary disks

### 2017-09-11 summary.tex
revising, new section 7.1

@@

@@

\subsection{Dusty analogs of other gaseous instabilities}

@@

\subsection{Implications for numerical simulations}
The dusty/adiabatic gas equivalence means that pure
gas dynamics codes can be applied to simulate locally
isothermal/polytropic dusty gas. One need not implement a separate 
equation for the dust component: 
%In fact, no
%modification is needed  
%for strictly isothermal or polytropic gas perfectly-coupled to dust:  
 the standard energy equation substitutes for the dust continuity
equation. When $c_s$ (or $K$) is not constant and/or $\tstop\neq0$ one
should add corresponding source terms in the energy equation
(e.g. Eq. \ref{poly_energy}). 
The source term associated with dust-gas drag, $\mathcal{C}$, is 
analogous to radiative diffusion \citep{price15} or thermal
conduction, which is also common in modern hydrodynamics codes.

@@

For the VSI we find dust-loading is generally stabilizing. In our disk models 
dust-loading does not affect VSI growth rates significantly, but
meridional motions may be suppressed where the dust-induced
vertical buoyancy dominates over vertical shear, consistent with our
previous study \citep{lin15}. 
Since the dust-induced buoyancy forces
increase away from the midplane, we find dust-loading can stabilize
`surface modes'  of the VSI, that would otherwise have the largest
growth rates. {\bf We also showed that radial variations in
  $\rhod/\rhog$ translates to an effectively non-uniform radial temperature
  profile, and the resulting vertical shear can lead to instability,
  even if the true sound-speed is constant. 
}

>>>>

{\bf

@@

\subsection{Applications and limitations}

The main application we envision for the dusty/adiabatic gas analogy 
to develop physical interpretations of dust-gas instabilities; and to find 
dusty analogs of pure gas instabilities in protoplanetary disks, 
which is discussed in \S\ref{dust_analogs}. One can also exploit the similarity to adapt existing
hydrodynamic codes to simulate dusty protoplanetary disks (see \S\ref{dust_sims}).

It is important to keep in mind the assumptions 
used to develop our thermodynamic model of dusty gas. 
The terminal velocity approximation, $\bm{v}_\mathrm{d} - \bm{v}_\mathrm{g} = \tstop \nabla P /\rhog$, was employed from the outset. 
%It assumes  
%the dust-gas relative drift $\bm{v}_mathrm{d} - \bm{v}_\mathrm{g} = \tstop \nabla P /\rhog$ 
This is applicable to small particles with 
$\tstop\OmK\ll 1$, but the precise critical value is problem-dependent. For the linear streaming instability,
even $\tstop\OmK \sim 0.1$ can be captured by simplified one-fluid equations 
(Fig. \ref{si_compare_fig}). On the other hand, the simplified equations contain spurious modes that
may overwhelm physical instabilities in certain situations (see \S\ref{spurious_epi}).  %Ultimately, one must check against the full equations.

However, our thermodynamic interpretation of dust-gas drift does \emph{not} in fact depend on the terminal velocity approximation.  
Once the gas equation of state is fixed and the true energy equation deleted, 
the dust continuity equation can be converted to a new effective energy equation. 
%This source term reflects  
%a heat flux proportional to $\bm{v}_\mathrm{d} - \bm{v}_\mathrm{g}$, which must generally be solved via a separate 
%evolutionary equation \citep{laibe14}. Nevertheless, this does not change the interpretation of dusty gas as a single ideal fluid subject to cooling.  
As an example, for strictly isothermal dusty gas we obtain   
\begin{align}
\frac{DP}{Dt} = - P\nabla\cdot\bm{v} + c_s^2\nabla\cdot\left[\fdust\left(1-\fdust\right)\rho\left(\bm{v}_\mathrm{d} - \bm{v}_\mathrm{g}\right)\right], \label{no_term_vel} 
\end{align}
\citep[see also][]{laibe14}. This equation does not use the terminal velocity approximation (which would give Eq. \ref{eff_energy}), but 
 we see that dust-gas relative drift can be regarded as a heat flux within the mixture. Calculating this heat flux generally requires solving the evolutionary equation for 
$\bm{v}_\mathrm{d} - \bm{v}_\mathrm{g}$. Nevertheless, one can still consider the dusty-gas as a single fluid subject to cooling. 
}

@@

\subsection{Dusty analogs of other gaseous instabilities} \label{dust_analogs}

@@

\subsection{Implications for numerical simulations}\label{dust_sims}
The dusty/adiabatic gas equivalence means that pure
gas dynamics codes can be applied to simulate locally
isothermal/polytropic dusty gas. One need not implement a separate 
equation for the dust component: 
%In fact, no
%modification is needed  
%for strictly isothermal or polytropic gas perfectly-coupled to dust:  
 the standard energy equation substitutes for the dust continuity
equation. When $c_s$ (or $K$) is not constant and/or $\tstop\neq0$ one
should add corresponding source terms in the energy equation
(e.g. Eq. \ref{poly_energy}). 
The source term associated with dust-gas drag, $\mathcal{C}$, is 
analogous to radiative diffusion \citep{price15} or thermal
conduction, which is also common in modern hydrodynamics codes.

@@

For the VSI we find dust-loading is generally stabilizing. In our disk models 
dust-loading does not affect VSI growth rates significantly, but
meridional motions may be suppressed where the dust-induced
vertical buoyancy dominates over vertical shear, consistent with our
previous study \citep{lin15}. 
Since the dust-induced buoyancy forces
increase away from the midplane, we find dust-loading can stabilize
`surface modes'  of the VSI, that would otherwise have the largest
growth rates. {\bf We also show that radial variations in
  $\rhod/\rhog$ effectively translates to a non-uniform radial temperature
  profile, and the resulting vertical shear can lead to instability,
  even if the true sound-speed is constant. 
}
